---
title: "02 - Get"
author: Jeffrey W. Hollister
layout: post_page
---

```{r setup, echo=FALSE, warning=FALSE}
options(repos="http://cran.rstudio.com/")
```

The second lesson is going to start to lay the foundation for working with data in R.  We will cover some of the very basics of R first, then move on to how you get data into R and how you work with some of the basic data structures.  Lastly, we will cover some ways to find relevant data and pull it directly into R.

##Quick Links to Exercises and R code
- [Lesson 2 R Code](http://jwhollister.com/gedr/rmd_posts/2015-01-14-02-Get.R): All the code from this post in an R Script.
- [Exercise 1](#exercise-1): Introduction to the console and basic functions

##Lesson Goals
- Understand workflow basics
- Learn R concepts such as assignment and operators
- Learn some useful R functions
- Understand the basic data structures and data types
- Be able to read data into R from a variety of sources

##More basics
Before we jump into details on data in R we need to introduce a few of the basics about working in the console, working via scripts, and workspaces/projects.  We will also start working with some simple, yet important R functions.

###Workflow
Being thoughtful about workflow from the beginning of a project is something that gets overlooked a lot, but a little up-front effort can provide a big benefit.  For our purposes we are going to make use of RStudio projects and script **EVERYTHING**.  The console has its utility and we will use it plenty, especially when we are first figuring out how to use commands.  But beyond that, we will store all of our work in a script.  The basic workflow I am going to advocate is:

1. Use a single project for this workshop (we created that in Lesson 1: Exercise 1).
2. Start everything in a script and use copy/paste or the RStudio tools to send to the console.
3. Use a new script for each lesson.
4. Comment ruthlessly.
5. Don't save .RData files or workspace history.  The script should recreate whatever you need.

This, of course, is not the only way you can strucutre a workflow, but I think it should be a good starting point for this workshop and one you can adapt to you own work afterwards.  And to provide a little motiviation, the more you utilize scripts, the more reproducible you work is, the more likely you will be able to recall what you did 3 months from now (your future self will thank you), and the easier it will be to transition your work from scripts to functions and, ultimately, to R packages.

###Working in the Console
As I mentioned above, the conosle and using R interactively is very powerful.  We will do this quite a bit.  Let's spend a little time playing around in the console and learn a few new functions.

R can be used as a caculator and a way to compare values.  Some examples of the basic operators:

```{r operators_consoloe}
#A really powerful calculator!
1+1 #Add
10-4 #Subtract
3*2 #Multiply
3^3 #Exponents
100/10 #Divide
5%%2 #Modulus
5>2 #Greater than
4<5 #Less than
5<=5 #Less than or equal
8>=2 #Greater than or equal
```

That's neat, but so what...  

Well, it could be interesting to do something with those values and save them for re-use.  We can do that with objects (everything in R is an object) and use the assignment operator, `<-`.  Know that object names cannot start with a number, contain spaces, or (most) special characters.  Underscore and periods are allowed.

```{r assignment_operator}
#Numeric assignment
x<-5
x
y<-x+1
y
z<-x+y
z
#Character
a<-"Bob"
a
b<-"Sue"
b
```

Now that we have a little experience working in the console and creating objects with `<-`, we might want to be able to do some additional things to navigate around, look at these objects etc.

Some functions that you might fund useful for working in with your R workspace:

```{r useful_functions_workspace}
#List all objects in current workspace
ls() 
ls(pattern="a")

#Remove an object
rm(x)

#Save your workspace
#Saves the whole thing to a file called lesson2.RData
save.image("lesson2.RData") 
#Saves just the a and y objects to a file called lesson2_ay.RData
save(a,y,file="lesson2_ay.RData")
```

This is probably a good spot to bring up quotes vs no quotes around arguments in a function.  This is a very common stumbling block.  The general rule is that no quotes are used only when referring to an object that currently exists.  Quotes are used in all other cases.  For instance in `save(a,y,file="lesson2_ay.RData")` the objects `a` and `y` are not quoted becuase they are objects in the workspace.  `file` is an argument of save and arguments are never quoted.  We quote the name of the file "lesson2_ay.RData" becuase it is not an R object but the name of a file to be created.  

You will likely still have some issues with this.  My recomendataion is to think about if it is an object in your R workspace or not.  If so, no quotes!

Next thing you might want to do is navigate around your files and directories.

```{r useful_functions_directory}
#See the current directory
getwd()

#Create a directory
dir.create("temp")

#Change the directory
setwd("temp")

#List files and directories
list.files()
list.files("..")
list.dirs("..")
```

##Exercise 1
For this first excercise I am actually going to ignore my workflow advice from above.  We are still in explore mode and saving this as a script doesn't yet make a whole lot of sense. Remember to use the green stickies when you have completed, and red stickies if you are running into problems. So, for this exercise:

1. Create two objects named `number1` and `number2` and give them the values of 25 and 4, respectively 
2. Create two more objects named `string1` and `string2`, give them any character string that you would like. 
3. Now using `number1`,`number2`, and the power of math create an object called `number3` that equals 100.
4. Create two more objects whose value is of your choosing
5. List the objects in your workspace
6. Remove `string2`
6. List your working directory
7. List the files in your working directory
8. Make sure your working directory is "gedr_workshop", if not use `setwd()` to get there.
9. Create a new directory called "R"
10. Try to add `string1` and `number1`.  What happens?  

So the last question in exercise 1 was a bit of a contrived way to segue into data types and structures.  So with that last bit, what did we ask R to do?  Why did it respond the way it did?  

In short it has a lot to do with data types.  Let's learn some more

##Data types and data structures in R
*Borrowed liberally from Jenny Bryan's [course materials on r](http://www.stat.ubc.ca/~jenny/STAT545A/quick-index.html) and Karthik Ram's [material from the Canberra Software Carpentry R Bootcamp](https://github.com/swcarpentry/2013-10-09-canberra).  Anything good is because of Jenny and Karthik.  Mistakes are all mine.*

Remember that everything in R is an object.  With regards to data, those objects have some specific characteristics that help R (and us) know what kind of data we are dealing with and what kind of operations can be done on that data.  This stuff may be a bit dry, but a basic understanding will help as so much of what we do with analysis has to do with the organization and type of data we have. First, lets discuss the atomic data types.

###Data Types
There are 5 basic atomic classes: character, numeric (real or decimal), integer, logical, and complex.

| Example | Type |
| ------- | ---- |
| "a", "swc" | character |
| 2, 15.5 | numeric | 
| 2 (usually add a `L` at end to denote integer)      | integer |
| `TRUE`, `FALSE` | logical |
| 1+4i | complex | 

In this workshop we will deal almost exclusively with three (and these are, in my experience, by far the most common): character, numeric, and logical.  There are also some useful commands to explore these data types and get some basic information about them.

``` {r data_type_functions, eval=FALSE}
typeof()  # what is it?
length()  # how long is it? What about two dimensional objects?
attributes()  # does it have any metadata?
```

###Data Structures
The next set of information relates to the many data strucutres in R.  Our efforts will focus on vectors and data frames.  We will discuss just the basics of lists and factors.  I will leave it to your curiousity to explore the matrix and table data structures

The data structres in base R include:

- vector
- list
- matrix
- data frame
- factors
- tables

### Vectors
A vector is the most common and basic data structure in `R` and is pretty much the workhorse of R. Vectors can be of two types:

* atomic vectors
* lists

**Atomic Vectors**
A vector can be a vector of characters, logical, integers or numeric.

Create an empty vector with `vector()`

```{r}
x <- vector()
x
# with a pre-defined length
x <- vector(length = 10)
x
# with a length and type
vector("character", length = 10)
vector("numeric", length = 10)
vector("integer", length = 10)
vector("logical", length = 10)
```

The general pattern is `vector(class of object, length)`.  You can also create vectors by concactenating them using the `c()` function.

Various examples:

```{r}
x <- c(1, 2, 3)
```

x is a numeric vector. These are the most common kind. They are numeric objects and are treated as double precision real numbers. To explicitly create integers, add a `L` at the end.

```{r}
x1 <- c(1L, 2L, 3L)
```

You can also have logical vectors. 

```{r}
y <- c(TRUE, TRUE, FALSE, FALSE)
```

Finally you can have character vectors:

```{r}
z <- c("Alec", "Dan", "Rob", "Karthik")
```

**Examine your vector**  

```{r}
typeof(z)
length(z)
class(z)
str(z)
```

Question: Do you see a property that's common to all these vectors above?

**Add elements**

```{r}
z <- c(z, "Annette")
z
```

More examples of vectors

```{r}
x <- c(0.5, 0.7)
x <- c(TRUE, FALSE)
x <- c(T, F)
x <- c("a", "b", "c", "d", "e")
x <- 9:100
x <- c(i + 0, 2 + 4)
```

You can also create vectors as sequence of numbers

```{r}
series <- 1:10
seq(10)
seq(1, 10, by = 0.1)
```

**Other objects**

`Inf` is infinity. You can have positive or negative infinity.

```{r}
1/0
# [1] Inf
1/Inf
# [1] 0
```

`NaN` means Not a number. it's an undefined value.

```{r}
0/0
NaN.
```


Each object has an attribute. Attribues can be part of an object of R. These include 

* names
* dimnames
* length
* class
* attributes (contain metadata)

For a vector, `length(vector_name)` is just the total number of elements.

### List

In R lists act as containers. Unlike atomic vectors, its contents are not restricted to a single mode and can encompass any data type. Lists are sometimes called recursive vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors. 

List is a special vector. Each element can be a different class.

Create lists using `list` or coerce other objects using `as.list()`

```{r}
x <- list(1, "a", TRUE, 1 + 4)
```

```{r}
x <- 1:10
x <- as.list(x)
length(x)
```


```{r}
xlist <- list(a = "Karthik Ram", b = 1:10, data = head(iris))
```

what is the length of this object?
what about its structure?

List can contain as many lists nested inside.

```{r}
temp <- list(list(list(list())))
temp
is.recursive(temp)
```

Lists are extremely useful inside functions. You can "staple" together lots of different kinds of results into a single object that a function can return.

It doesn't print out like a vector. Prints a new line for each element.

### Factors

Factors are special vectors that represent categorical data. Factors can be ordered or unordered and are important when for modelling functions such as `lm()` and `glm()` and also in plot methods.

Factors can only contain pre-defined values.

Factors are pretty much integers that have labels on them.  While factors look (and often behave) like character vectors, they are actually integers under the hood, and you need to be careful when treating them like strings. Some string methods will coerce factors to strings, while others will throw an error.

Sometimes factors can be left unordered. Example: male, female

Other times you might want factors to be ordered (or ranked). Example: low, medium, high. 

Underlying it's represented by numbers 1,2,3.

They are better than using simple integer labels because factors are what are called self describing. male and female is more descriptive than 1s and 2s. Helpful when there is no additional metadata.

Which is male? 1 or 2? You wouldn't be able to tell with just integer data. Factors have this information built in.

Factors can be created with `factor()`. Input is a character vector.

```{r}
x <- factor(c("yes", "no", "no", "yes", "yes"))
x
```

`table(x)` will return a frequency table.

`unclass(x)` strips out the class information.

In modeling functions, importnat to know whta baseline levels is.
This is the first factor but by default the ordering is determined by alphabetical order of words entered. You can change this by speciying the levels.

```{r}
x <- factor(c("yes", "no", "yes"), levels = c("yes", "no"))
```

### Data frame

A data frame is a very important data type in R. It's pretty much the de facto data structure for most tabular data and what we use for statistics.

data frames can have additional attributes such as `rownames()`. This can be useful for annotating data, like subject_id or sample_id. But most of the time they are not used.

e.g. `rownames()`
useful for annotating data. subject names.
other times they are not useful.

* Data frames Usually created by read.csv and read.table.

* Can convert to `matrix` with `data.matrix()`

* Coercion will force and not always what you expect.

* Can also create with `data.frame()` function.

With and data frame, you can do `nrow(df)` and `ncol(df)`
rownames are usually 1..n.

**Useful functions**  
`head()` - see first 5 rows
`tail()` - see last 5 rows
`dim()` - see dimensions
`nrow()` - number of rows
`ncol()` - number of columns
`str()` - structure of each column
`names()` - will list column names for a data.frame (or any object really).
`summary()` - summarizes each column in a data frame

##Reading data into R

##Exercise 2