---
title: "02 - Get"
author: Jeffrey W. Hollister
layout: post_page
---

```{r setup, echo=FALSE, warning=FALSE}
options(repos="http://cran.rstudio.com/")
```

The second lesson is going to start to lay the foundation for working with data in R.  We will cover some of the very basics of R first, then move on to how you get data into R and how you work with some of the basic data structures.  Lastly, we will cover some ways to find relevant data and pull it directly into R.

##Quick Links to Exercises and R code
- [Lesson 2 R Code](http://jwhollister.com/gedr/rmd_posts/2015-01-14-02-Get.R): All the code from this post in an R Script.

##Lesson Goals
- Understand workflow basics
- Learn R concepts such as assignment and operators
- Learn some useful R functions
- Understand the basic data structures and data types
- Be able to read data into R from a variety of sources
- Interact with data being served from web APIs

##More basics
Before we jump into details on data in R we need to introduce a few of the basics about working in the console, working via scripts, and workspaces/projects.  We will also start working with some simple, yet important R functions.

###Workflow
Being thoughtful about workflow from the beginning of a project is something that gets overlooked a lot, but a little up-front effort can provide a big benefit.  For our purposes we are going to make use of RStudio projects and script **EVERYTHING**.  The console has its utility and we will use it plenty, especially when we are first figuring out how to use commands.  But beyond that, we will store all of our work in a script.  The basic workflow I am going to advocate is:

1. Use a single project for this workshop (we created that in Lesson 1: Exercise 1).
2. Start everything in a script and use copy/paste or the RStudio tools to send to the console.
3. Use a new script for each lesson.
4. Comment ruthlessly.
5. Don't save .RData files or workspace history.  The script should recreate whatever you need.

This, of course, is not the only way you can strucutre a workflow, but I think it should be a good starting point for this workshop and one you can adapt to you own work.  And to provide a little motiviation, the more you utilize scripts, the more reproducible you work is, the more likely you will be able to recall what you did 3 months from now (your future self will thank you), and the easier it will be to transition your work from scripts to functions and, ultimately, to R packages.

###Working in the Console
As I mentioned above, the conosle and using R interactively is very powerful.  We will do this quite a bit.  Let's spend a little time playing around in the console and learn a few new functions.

R can be used as a caculator and a way to compare values.  Some examples of the basic operators:

```{r operators_consoloe}
#A really powerful calculator!
1+1 #Add
10-4 #Subtract
3*2 #Multiply
3^3 #Exponents
100/10 #Divide
5%%2 #Modulus
5>2 #Greater than
4<5 #Less than
5<=5 #Less than or equal
8>=2 #Greater than or equal
```

That's neat, but so what...  

Well, it could be interesting to do something with those values and save them for re-use.  We can do that with objects (everything in R is an object) and use the assignment operator, `<-`.

```{r assignment_operator}
#Numeric assignment
x<-5
x
y<-x+1
y
z<-x+y
#Character
a<-"Bob"
a
b<-"Sue"
b
```

##Exercise 1


##Data types and structures in R

##Reading data into R

##Where to find data