---
title: "02 - Get"
author: Jeffrey W. Hollister
layout: post_page
---

```{r setup, echo=FALSE, warning=FALSE}
options(repos="http://cran.rstudio.com/")
```

The second lesson is going to start to lay the foundation for working with data in R.  We will cover some of the very basics of R first, then move on to how you get data into R and how you work with some of the basic data structures.  Lastly, we will cover some ways to find relevant data and pull it directly into R.

##Quick Links to Exercises and R code
- [Lesson 2 R Code](http://jwhollister.com/gedr/rmd_posts/2015-01-14-02-Get.R): All the code from this post in an R Script.
- [Exercise 1](#exercise-1): Introduction to the console and basic functions

##Lesson Goals
- Understand workflow basics
- Learn R concepts such as assignment and operators
- Learn some useful R functions
- Understand the basic data structures and data types
- Be able to read data into R from a variety of sources

##More basics
Before we jump into details on data in R we need to introduce a few of the basics about working in the console, working via scripts, and workspaces/projects.  We will also start working with some simple, yet important R functions.

###Workflow
Being thoughtful about workflow from the beginning of a project is something that gets overlooked a lot, but a little up-front effort can provide a big benefit.  For our purposes we are going to make use of RStudio projects and script **EVERYTHING**.  The console has its utility and we will use it plenty, especially when we are first figuring out how to use commands.  But beyond that, we will store all of our work in a script.  The basic workflow I am going to advocate is:

1. Use a single project for this workshop (we created that in Lesson 1: Exercise 1).
2. Start everything in a script and use copy/paste or the RStudio tools to send to the console.
3. Use a new script for each lesson.
4. Comment ruthlessly.
5. Don't save .RData files or workspace history.  The script should recreate whatever you need.

This, of course, is not the only way you can strucutre a workflow, but I think it should be a good starting point for this workshop and one you can adapt to you own work afterwards.  And to provide a little motiviation, the more you utilize scripts, the more reproducible you work is, the more likely you will be able to recall what you did 3 months from now (your future self will thank you), and the easier it will be to transition your work from scripts to functions and, ultimately, to R packages.

###Working in the Console
As I mentioned above, the conosle and using R interactively is very powerful.  We will do this quite a bit.  Let's spend a little time playing around in the console and learn a few new functions.

R can be used as a caculator and a way to compare values.  Some examples of the basic operators:

```{r operators_consoloe}
#A really powerful calculator!
1+1 #Add
10-4 #Subtract
3*2 #Multiply
3^3 #Exponents
100/10 #Divide
5%%2 #Modulus
5>2 #Greater than
4<5 #Less than
5<=5 #Less than or equal
8>=2 #Greater than or equal
```

That's neat, but so what...  

Well, it could be interesting to do something with those values and save them for re-use.  We can do that with objects (everything in R is an object) and use the assignment operator, `<-`.  Know that object names cannot start with a number, contain spaces, or (most) special characters.  Underscore and periods are allowed.

```{r assignment_operator}
#Numeric assignment
x<-5
x
y<-x+1
y
z<-x+y
z
#Character
a<-"Bob"
a
b<-"Sue"
b
```

Now that we have a little experience working in the console and creating objects with `<-`, we might want to be able to do some additional things to navigate around, look at these objects etc.

Some functions that you might fund useful for working in with your R workspace:

```{r useful_functions_workspace}
#List all objects in current workspace
ls() 
ls(pattern="a")

#Remove an object
rm(x)

#Save your workspace
#Saves the whole thing to a file called lesson2.RData
save.image("lesson2.RData") 
#Saves just the a and y objects to a file called lesson2_ay.RData
save(a,y,file="lesson2_ay.RData")
```

This is probably a good spot to bring up quotes vs no quotes around arguments in a function.  This is a very common stumbling block.  The general rule is that no quotes are used only when referring to an object that currently exists.  Quotes are used in all other cases.  For instance in `save(a,y,file="lesson2_ay.RData")` the objects `a` and `y` are not quoted becuase they are objects in the workspace.  `file` is an argument of save and arguments are never quoted.  We quote the name of the file "lesson2_ay.RData" becuase it is not an R object but the name of a file to be created.  

You will likely still have some issues with this.  My recomendataion is to think about if it is an object in your R workspace or not.  If so, no quotes!

Next thing you might want to do is navigate around your files and directories.

```{r useful_functions_directory}
#See the current directory
getwd()

#Create a directory
dir.create("temp")

#Change the directory
setwd("temp")

#List files and directories
list.files()
list.files("..")
list.dirs("..")
```

##Exercise 1
For this first excercise I am actually going to ignore my workflow advice from above.  We are still in explore mode and saving this as a script doesn't yet make a whole lot of sense. Remember to use the green stickies when you have completed, and red stickies if you are running into problems. So, for this exercise:

1. Create two objects named `number1` and `number2` and give them the values of 25 and 4, respectively 
2. Create two more objects named `string1` and `string2`, give them any character string that you would like. 
3. Now using `number1`,`number2`, and the power of math create an object called `number3` that equals 100.
4. Create two more objects whose value is of your choosing
5. List the objects in your workspace
6. Remove `string2`
6. List your working directory
7. List the files in your working directory
8. Make sure your working directory is "gedr_workshop", if not use `setwd()` to get there.
9. Create a new directory called "R"
10. Try to add `string1` and `number1`.  What happens?  

So the last question in exercise 1 was a bit of a contrived way to segue into data types and structures.  So with that last bit, what did we ask R to do?  Why did it respond the way it did?  

In short it has a lot to do with data types.  Let's learn some more

##Data types and data structures in R
*Borrowed liberally from Jenny Bryan's [course materials on r](http://www.stat.ubc.ca/~jenny/STAT545A/quick-index.html) and Karthik Ram's [material from the Canberra Software Carpentry R Bootcamp](https://github.com/swcarpentry/2013-10-09-canberra).  Anything good is because of Jenny and Karthik.  Mistakes are all mine.*

Remember that everything in R is an object.  With regards to data, those objects have some specific characteristics that help R (and us) know what kind of data we are dealing with and what kind of operations can be done on that data.  This stuff may be a bit dry, but a basic understanding will help as so much of what we do with analysis has to do with the organization and type of data we have. First, lets discuss the atomic data types.


###Data Types
There are 5 basic atomic classes: character, numeric (real or decimal), integer, logical, and complex.

| Example | Type |
| ------- | ---- |
| "a", "swc" | character |
| 2, 15.5 | numeric | 
| 2 (usually add a `L` at end to denote integer)      | integer |
| `TRUE`, `FALSE` | logical |
| 1+4i | complex | 

In this course we will deal almost exclusively with three (and these are, in my experience, by far the most common): character, numeric, and logical.

##Reading data into R

##Exercise 2